# Getting Started

## Prerequisites

Basic knowledge of working with datasets in R is essential. This course assumes that you're comfortable with reading datasets, working with script files, and navigating in RStudio.

## Software Requirements

### R and RStudio

Recent versions of R (version 3.2 or newer) and RStudio (version 1.0 above) are required.

You can download the latest versions from the links below:

- [Download R](https://cran.r-project.org)
- [Download RStudio](https://www.rstudio.com/products/rstudio/download)

You can find out the version of R installed by typing `version` at the console:

```{r}
version
```

## Required Packages

This workshop relies on three packages: `dplyr`, `tidyr`, and `readr`. There are two ways to install these packages:

### Option 1: Use `tidyverse`

You can either install these two packages individually or use `tidyverse`. The `tidyverse` package is a collection of packages used for data manipulation and visualization. In addition to `dplyr`, `tidyr`, and `readr`, it also includes the following:

```{r echo = FALSE}
tidyverse::tidyverse_packages()
```

You can install `tidyverse` using the `install.packages()` function:

```{r eval = FALSE}
install.packages("tidyverse")
```

You can find out the version of `tidyverse` installed using the `packageVersion()` function:

```{r}
packageVersion("tidyverse")
```

To update `tidyverse` packages, you can use the `tidyverse_update()` function:

```{r eval = FALSE}
tidyverse::tidyverse_update()
```

### Option 2: Install Individual Packages

If you encounter any problems installing `tidyverse`, then the other option is to install `dplyr`, `tidyr`, and `readr` individually.

```{r eval = FALSE}
install.packages("dplyr")
install.packages("tidyr")
install.packages("readr")
```

